---
title: "Modeling Microbial Mutualisms"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## Overview
Simulation model that explores interactions between Esherichia coli and Rhodopseudomonas palustris. Under anaerobic conditions, glucose is consumed by E. coli and converts some of this to fermentation byproducts, which serve as a carbon and energy source for R. palustris. In turn, R. palustris fixes atmospheric nitrogen, which is converted into ammonia and serves as E. coli's sole source of nitrogen. Following attempts to captures these baisc features using a set of coupled differential equations. 

Modified by Jake for batch growth

## Set working directory
```{r}
rm(list=ls()) # clears your working directory
getwd() # tells you your current working directory
setwd("~/GitHub/Task3") # sets your working directory
```

## Install packages
```{r}
# install.packages("deSolve") # run this first time
require("deSolve") # installs any dependencies
```

## Mutualism function
```{r}
mutfun <- function(times, y, parms) {
  with(as.list(c(y, parms)), {
   ue <- uE * (G/(Kg + G)) * ((A + Ax) /(Kae + A + Ax)) * (be/(be+C+Fr))
   ur <- uR * (N/(Kn + N)) * ((C + Cx)/(Kc + C + Cx)) * (br/(br+C+Fr)) # + (uR2 * ((A+Ax) /(Kar+A+Ax)) * ((C+Cx)/(Kc + C + Cx)))
    #dgdt <- (D * Gin) - (ue*E*a) - (E*a*(G/(Kg+G))) - (ue*E/Yg) - (D * G) # Glucose
    #dndt <- (D * Nin) - (ur*R*z) - (ur * R / Yn) - (D * N) # N2
  #  dcdt <- (E*ue*a) + (E*a*(G/(Kg + G))) - (ur*R/Yc) - (D*C) # OAcs
  #  dadt <- (R*ur*z)  - (ue*E/Ya) - (D * A) # NH4
  #  dedt <- ue * E - (D * E) # E. coli
  #  drdt <- ur * R - (D * R) # R. palustris
    
    dgdt <- -ue*E/Yg  - (ue*E*(c+f+e)) - E*(c+f+e)*(G/(Kg+G))*(ng/(ng+ue))#Glucose
    dndt <- - (ur*R*z) - (ur * R / Yn)  # N2
   dcdt <- (E*ue*c) + (E*c*(G/(Kg + G))*(ng/(ng+ue))) - (ur*R/Yc)# - (D*C)# OAcs except formate
   dfdt <- (E*ue*f) + (E*f*(G/(Kg+G))*(ng/(ng+ue))) #formate excretion rate
    # dcdt <- (E*ue*a) + (E*a*(G/(Kg + G))) - (ur*R/Yc) - (D*C)# OAcs
    dadt <- (R*ur*z)  - (ue*E/Ya) #- (ur*R/Yar)# NH4
    
    dedt <- ue * E  # E. coli
    drdt <- ur * R   # R. palustris
    dEtdt <- (E*ue*e) + (E*e*(G/(Kg+G))*(ng/(ng+ue)))
    return(list(c(dgdt, dndt, dcdt, dfdt, dadt, dedt, drdt,dEtdt))) #dcdt, dndt, dadt, drdt
  })
}
```

## Parameters and initial conditions
```{r}
# Parameters
parms <- c(uE = 0.347, # E. coli max growth rate (h^-1)
           uR = 0.063, # R. palustrix max growth rate (h^-1)
           uR2 = 0.0236, # OPTION represents boost to new max in presence of NH4+ (h^-1)
           #ue and ur are Rpal growth rates at any given time
           Yg = 20000000000,#13888800000, # Ec growth yield cells/mmol each Glc C 
           Yn = 94423800000, # Rp growth yield cells/mmol each N2 N
           Yc = 82994820000, #Rp growth yield on org acids - based on cells/mmol ace C 
           Ya = 129581200000, #Ec growth yield on on NH4+ (cells/mmol)
           Yar = 259000000, #OPTION - needs updated value - Rp growth yield on on NH4+ (cells/mmol)
        D = 0.26,#/48, # CSTAT OPTION Dilution rate (h^1)
           Kg = 0.020, # half saturation constant for glucose (mM) Buhr et al 1992 J Biol Chem 267:3847-51
           Kc = 0.011, # half saturation constatnt for fermentation product (mM) (assumed)
           Kn = 0.100, # half saturation constant for N2 (mM) PMID: 440158
           Kae = 0.010, # half saturation constant for ammonia (mM) by Ec http://www.sciencemag.org/content/305/5690/1587
          Kar = 0.015, # OPTION half saturation constant for ammonia (mM) by Rp
           a = 1.625*10^-7, # OPTION conversion for G -> all ferment prod per E. coli
           c = 4*10^-11,#7.14*10^-11, # mmol per CFU - all products except formate
           f = 1.08*10^-10, # mmol formate per CFU
           e = 2.66*10^-11, # mmol etoh per CFU
           z = 1*10^-13,#, # conversion for N2 -> NH3 per R. palustris (assumed)
          ng = 0.01, # factor used in formula to shift between growing vs nongrowing C-excretion
        Gin <- 25, #CSTAT OPTION
         Nin <- 100, #CSTAT OPTION
        be<-20, # resistance of Ecoli to acid; reflects medium buffering
        br<-1, # resistance of Rpal to acid; reflects medium buffering
        Ax<-0, #OPTION externally added ammonium
        Cx<-0) #OPTION externally added Oacs           
# Initial conditions  THE ORDER HERE MUST MATCH THE RETURNS ABOVE 
init <- c(G = 25,# Glucose (mM)
          N = 76,    # N2 (mM) 
         C = 0, # Fermentation product except for (mM)
         Fr = 0, #formate excreted
          A = 0, # NH4 (mM)
          E = 10000000,      # E. coli (biomass? units?)
          R = 10000000,     # R. palustris (biomass?, units?)
          Et = 0) #ethanol
# Time range and time step
Tmax = 3200
TimeStep = 1
Time = seq(0, Tmax, by = TimeStep)
```

## Output
```{r}
out <- ode(y = init, times = Time, func = mutfun, parms = parms, method = "lsode")
```

## Plotting
```{r}
plot(out[,1], out[,7], xlab = "time (hr)", ylab = "Ecoli", type = "l", las = 1, 
     col = "gold", lwd = 4, ylim = c(0, 1*max(out[,7]))) # plot E. coli biomass
par(new=T)
plot(out[,1], out[,8], xlab="", ylab ="",  type = "l",las = 1,
     col = "purple", ylim=c(0,1.2*max(out[,8])) )
axis(4, ylim=c(0,1.2*max(out[,7])), lwd=2, line=3.5)
mtext(4, text="Rpal", line=5.5)
par(new=T)
plot(out[,1], out[,4], xlab="", ylab ="",  type = "l",las = 1,
     col = "green", ylim=c(0,1.2*max(out[,4])) )
axis(4, ylim=c(0,1.2*max(out[,4])), lwd=2, line=3.5)
mtext(4, text="Oacs", line=5.5)
par(new=T)

plot(out[,1], out[,6], xlab="", ylab ="",  type = "l",las = 1,
     col = "blue", ylim=c(0,1.3*max(out[,6])) )
axis(4, ylim=c(0,0.1*max(out[,4])), lwd=2, line=3.5)
mtext(4, text="NH4+", line=5.5)
par(new=T)
plot(out[,1], out[,2], xlab="", ylab ="",  type = "l",las = 1,
     col = "black", ylim=c(0,1.4*max(out[,2])) )
axis(4, ylim=c(0,1.2*max(out[,2])), lwd=2, line=3.5)
mtext(4, text="glucose", line=5.5)

par(new=T)
plot(out[,1], out[,3], xlab="", ylab ="",  type = "l",las = 1,
     col = "gray", ylim=c(0,1.4*max(out[,3])) )
axis(4, ylim=c(0,1.2*max(out[,3])), lwd=2, line=3.5)
mtext(4, text="glucose", line=5.5)

#points(out[,1], out[,7], type = "l", col = "purple", lwd = 4)
#points(out[,1], out[,2], type = "l", col = "green", lwd = 4)
```